---
title: 'Day 1 Workshop: Essential Tools Setup'
subtitle: 'Git, GitHub, and R Markdown'
output:
  html_document: default
  pdf_document: default
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# This workshop doesn't require any special R packages for the setup portion
```

## Introduction

This workshop will help you set up the essential tools you'll need throughout this course. We'll cover:

1. **Git** - Version control for tracking changes in your code
2. **GitHub** - Online platform for hosting and sharing code
3. **R Markdown** - Document format for reproducible research
4. **Ollama**: Ollama is software that allows you to install and run large language models on your local machine.

By the end of this workshop, you'll have:
- Git installed and configured on your computer
- A GitHub account and repository for your project
- Basic understanding of how to use these tools together

## Prerequisites

Before we begin, make sure you have:
- A computer with internet access
- R and RStudio installed (if not, we'll help you install them)
- A GitHub account (we'll create one if needed)

## Part 1: Installing Git

### For Windows Users

1. **Download Git for Windows:**
   - Go to https://git-scm.com/download/win
   - Download the latest version for Windows
   - Run the installer

2. **Installation Options (recommended settings):**
   - Checkout as-is, commit Unix-style line endings
   - Default (fast-forward or merge)

3. **Verify Installation:**
   - Open Command Prompt or Git Bash
   - Type: `git --version`
   - You should see something like: `git version 2.40.0.windows.1`

### For Mac Users

1. **Using Homebrew (recommended):**
   ```bash
   # Install Homebrew first (if you don't have it)
   /bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"
   
   # Install Git
   brew install git
   ```

2. **Or download from Git website:**
   - Go to https://git-scm.com/download/mac
   - Download and install

3. **Verify Installation:**
   - Open Terminal
   - Type: `git --version`
   - You should see something like: `git version 2.40.0`

### Configure Git

After installation, configure Git with your information:

```bash
git config --global user.name "Your Name"
git config --global user.email "your.email@example.com"
```

## Part 2: Creating a GitHub Account

1. **Go to GitHub:**
   - Visit https://github.com
   - Click "Sign up"

2. **Create Account:**
   - Enter your email address
   - Create a password
   - Choose a username (this will be your GitHub handle)
   - Complete the verification process

3. **Add an SSH key to your GitHub account** (optional)
   - Adding an SSH key allows you to avoid endlessly re-typing your password when working with git.
   - Follow the instructions here: https://docs.github.com/en/authentication/connecting-to-github-with-ssh/adding-a-new-ssh-key-to-your-github-account


## Part 3: Setting Up Your Project Repository

### Create a New Repository

1. **On GitHub:**
   - Click the "+" icon in the top right
   - Select "New repository"

2. **Repository Settings:**
   - **Repository name:** `bst209-final-project` (or your preferred name)
   - **Description:** "Machine Learning in Healthcare Final Project"
   - **Visibility:** Public
   - **Initialize with:** Check "Add a README file"
   - Click "Create repository"

### Clone Your Repository

1. **Get the repository URL:**
   - On your GitHub repository page, click the green "Code" button
   - Copy the "SSH" URL if possible, if you set up your SSH key. If not, copy the HTTPS URL (looks like: `https://github.com/yourusername/bst209-final-project.git`)

2. **Clone to your computer:**
   - Open your shell (Terminal/GitBash)

   ```bash
   # Navigate to where you want your project
   cd ~/Documents  # or wherever you prefer
   
   # Clone the repository
   git clone <YOUR_REPO_LINK_COPIED_FROM_GITHUB>
   
   # Navigate into your project folder
   cd bst209-final-project
   ```

## Part 4: Basic Git Commands

Let's practice with some basic Git commands:

### Check Status
```bash
git status
```
This shows you the current state of your repository.

### Make Changes and Commit
```bash
# Create a new file
echo "# My BST209 Project" > README.md

# Check status
git status

# Add the file to staging
git add README.md

# Commit the changes
git commit -m "Initial commit with README"

# Push to GitHub
git push origin main
```

### Basic Workflow
Here's the typical workflow you'll use:

1. **Make changes** to your files
2. **Check status:** `git status`
3. **Add files:** `git add filename`
4. **Commit changes:** `git commit -m "Description of changes"`
5. **Push to GitHub:** `git push origin main`

## Part 5: R Markdown Basics

R Markdown combines R code, output, and text in a single document. It's perfect for reproducible research.

### Create Your First R Markdown Document

1. **In RStudio:**
   - File → New File → R Markdown
   - Give it a title: "BST209 Project"
   - Choose HTML output
   - Click "OK"

2. **Basic R Markdown Structure:**
   ```yaml
   ---
   title: "BST209 Project"
   author: "Your Name"
   date: "`r Sys.Date()`"
   output: html_document
   ---
   ```

3. **Add Content:**

   # Introduction
   
   This is my final project for BST209: Machine Learning in Healthcare.
   
   ## Data Analysis
   
   ```{r, eval=FALSE}
   # Your R code goes here
   library(tidyverse)
   
   # Load data
   data <- read.csv("your_data.csv")
   
   # Basic summary
   summary(data)
   ```

### Key R Markdown Features

- **Headers:** Use `#` for different levels
- **Code chunks:** Use backticks `` for R code
- **Lists:** Use `-` or `*` for bullet points
- **Bold:** Use `**text**`
- **Italic:** Use `*text*`

### Knitting Your Document

1. **Click the "Knit" button** in RStudio
2. **Choose output format** (HTML, PDF, Word)
3. **Your document will be rendered** with code output and formatting

## Part 6: Integrating Git with RStudio

RStudio has built-in Git support:

1. **Enable Git in RStudio:**
   - Tools → Global Options → Git/SVN
   - Set path to Git executable
   - Restart RStudio

2. **Using Git in RStudio:**
   - **Git tab:** Shows file status
   - **Commit button:** Stage and commit changes
   - **Push/Pull buttons:** Sync with GitHub

## Exercise: Create Your First Project File

Let's create your first project file and practice the workflow:

1. **Create an R Markdown file:**
   ```r
   # In RStudio, create a new R Markdown file
   # Save it as "project_overview.Rmd"
   ```

2. **Add some content:**

   ```markdown
   ---
   title: "BST209 Final Project Overview"
   author: "Your Name"
   date: "`r Sys.Date()`"
   output: html_document
   ---
   
   # Project Overview
   
   ## Research Question
   
   [Describe your research question here]
   
   ## Data Source
   
   [Describe your data source here]
   
   ## Methods
   
   [Describe your planned methods here]
   
   ## Expected Outcomes
   
   [Describe what you expect to find]
   ```

3. **Practice the Git workflow:**
   ```bash
   # Add your new file
   git add project_overview.Rmd
   
   # Commit the changes
   git commit -m "Add project overview document"
   
   # Push to GitHub
   git push origin main
   ```

## Part 7: Install Ollama

Ollama is software that allows you to install and run large language models on your local machine.

For the workshop on large language models, you will need to install Ollama.

To install Ollama, go to: https://ollama.com/

## Troubleshooting Common Issues

### Git Issues

**"Git is not recognized as a command":**
- Make sure Git is installed and in your PATH
- Restart your terminal/command prompt

### R Markdown Issues

**"Package not found":**
```{r, eval=FALSE}
# Install missing packages
install.packages("package_name")
```

## Best Practices

### Git Best Practices
- **Commit frequently** with meaningful messages
- **Use descriptive commit messages**
- **Pull before pushing** to avoid conflicts
- **Don't commit large files** (use .gitignore)

### R Markdown Best Practices
- **Keep code chunks focused** on one task
- **Use meaningful chunk names**
- **Document your code** with comments
- **Test your document** regularly

### Project Organization
```
bst209-final-project/
├── README.md
├── data/
├── code/
├── results/
├── docs/
└── .gitignore
```

## Next Steps

1. **Complete the setup** if you haven't already
2. **Practice the basic Git workflow** with small changes
3. **Create your project structure** following the organization above
4. **Start working on your project overview** document

## Resources

- **Git Documentation:** https://git-scm.com/doc
- **GitHub Guides:** https://guides.github.com/
- **R Markdown:** https://rmarkdown.rstudio.com/
- **RStudio Cheat Sheets:** https://www.rstudio.com/resources/cheatsheets/

---

**Remember:** The goal is to get comfortable with these tools, so use the project as an opportunity to practice.